\texttt{ensmallen}: a generic C++ library for fast optimization

### 2018-10-14 doc/mlsys_paper/paper.tex
Significant paper refactoring---fit into the required space.
Only a few TODOs left.

\documentclass{article}
\usepackage{nips_2018}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage[newfloat]{minted}
\usepackage{caption}
\usepackage{parcolumns}
\usepackage{booktabs}

@@

\newcolumntype{R}[2]{%
  >{\adjustbox{angle=#1,lap=\width-(#2)}\bgroup}%
  l%
  <{\egroup}%
}
\newcommand*\rot{\multicolumn{1}{R{30}{2em}}}

@@

% Alphabetical ordering?
\author{Shikhar Bhardwaj \And Ryan R. Curtin \And Marcus Edel \And Yannis
Mentekidis \And Conrad Sanderson}

@@

\begin{abstract}

%% the abstract below still needs more meat and sharpening

We present a fast and flexible C++ library for mathematical optimization of 
arbitrary user-supplied functions,
which can be applied to many machine learning problems.
The library provides many pre-built optimizers
(including numerous variants of SGD and Quasi-Newton optimizers)
as well as a flexible framework for implementing new optimizers and objective functions.
Implementation of a new optimizer requires only one method
and implementation of a new objective function requires at most four C++ functions. 
This can aid in the  quick implementation and prototyping of new machine learning algorithms.
In contrast to optimization approaches in existing frameworks
(such as Julia, Shogun, Vowpal Wabbit, SciPy),
the proposed Ensmallen library supports more optimizations,
more types of objective functions,
as well as seamlessly supporting user-defined objective functions and optimizers.
Empirical comparisons show that ...
TODO: briefly describe the speedups obtained in the experiments

\end{abstract}

@@

Mathematical optimization is the workhorse of virtually all machine learning
algorithms.  For a given objective function $f(\cdot)$
(which may have a special structure or constraints),
almost all machine learning problems can be boiled down
to the following optimization form:
%
\begin{equation}
\operatorname{argmin} f(x)
\end{equation}

@@

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms.
We present {\tt ensmallen}, a C++ optimization toolkit
that contains a wide variety of optimization techniques for many types of
objective functions.  Through the use of C++ template metaprogramming~\cite{TODO},
{\tt ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

Although there are many existing machine learning optimization toolkits, few
are able to take explicit advantage of metaprogramming based code optimizations,
and few offer robust support for various types of objective functions.
For instance, deep learning
libraries like Caffe~\cite{jia2014caffe},
PyTorch~\cite{paszke2017automatic},
and TensorFlow~\cite{abadi2016tensorflow}
each contain a variety of optimization techniques.  However, these techniques are
limited to stochastic gradient descent (SGD) and SGD-like optimizers that
operate on small batches of data points at a time.  Other machine learning
libraries, such as {\tt scikit-learn}~\cite{pedregosa2011scikit}
% TODO: can we find another library to put here too?  Shark maybe?
contain optimization algorithms but not in a coherent or reusable framework.
Many programming languages have higher-level packages for
mathematical optimization.  For example, {\tt scipy.optimize} % TODO: check and cite
is widely used in the Python community, and MATLAB's function optimization
support has been available and used for many decades.  However, these
implementations are often unsuitable for modern machine learning tasks---for
instance, computing the full gradient of the objective function may not be
feasible because there are too many data points.

@@

% Originally, {\tt ensmallen} was developed as part of {\tt mlpack}; % TODO: cite
% it is now available in standalone header-only form.
% TODO: seems awkward to put the lineage here;
%% CS: agreed; i don't see a need to explicitly mention mlpack -- it distracts
%% CS: from the central message
% TODO: also, we should talk up the use

\section{Types of objective functions}

In short, {\tt ensmallen} provides a {\bf set of optimizers} that can be used to
{\bf optimize user-defined objective functions}.  It's also easy to implement a
new optimizer in the {\tt ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem

\begin{equation}
\operatorname{argmin}_{x} f(x)
\end{equation}

\noindent for any function $f(x)$ that takes some vector or matrix input $x$.  Ideally,
the user should only need to specify what $f(x)$ is (or provide an
implementation of a function that can compute $f(x)$).  But it is impossible to
implement something both so generic and fast---for instance, gradient-based
approaches converge far more quickly than gradient-free approaches (in general),
so we must design an abstraction that is able to simultaneously generalize to
many problem types, as well as take advantage of accelerations and
optimizations.

\begin{itemize}
  \item {\bf arbitrary}: no assumptions can be made on $f(x)$
  \item {\bf differentiable}: $f(x)$ has a computable gradient $f'(x)$
  \item {\bf separable}: $f(x)$ is a sum of individual components: $f(x) =
\sum_{i} f_i(x)$
  \item {\bf categorical}: $x$ contains elements that can only take discrete
values
  %\item {\bf numeric}: all elements of $x$ take values in $\mathcal{R}$
  \item {\bf sparse}: the gradient $f'(x)$ or $f_i(x)$ (for a separable
function) is sparse
  \item {\bf partially differentiable}: the gradient $f'_j(x)$ is computable for
individual elements $x_j$ of $x$
  \item {\bf bounded}: $x$ is limited in the values that it can take
\end{itemize}

% Just to get a feeling of the look; Also, we could also just use \OK instead of
% \CIRCLE or \LEFTcircle
\begin{table}

@@

        & & \multicolumn{7}{c}{} \\[2ex]

@@

\caption{

@@

{\it unified framework} the library/language has some kind of generic and unified

@@

{\it categorical} refers to whether support for categorical features exists.
}

@@

@@

@@

For the most common case of a differentiable objective function, the user must
implement only two methods:

\begin{itemize}
  \item \texttt{double Evaluate($x$)}: given coordinates $x$, this function
should return the value of $f(x)$.
  \item \texttt{void Gradient($x$, $g$)}: given coordinates $x$ and a reference
to $g$, set $g = f'(x)$.
\end{itemize}

@@

\section{Example}

As an example of usage, let us consider the linear regression objective
function\footnote{Just for simplicity, we ignore the bias term.  It can be
rederived by taking $x^*_i = (x_i, 1)$.}.  Given some dataset $X \in
\mathcal{R}^{n \times d}$ and associated responses $y \in \mathcal{R}^n$, the
model of linear regression is to assume that $y_i = x_i \theta$ for each data
point and response $(x_i, y_i)$.  To fit this model $\theta \in \mathcal{R}^d$ best
to the data, we wish to minimize

\begin{equation}
\operatorname{argmin}_\theta \sum_{i = 1}^n (y_i - x_i \theta)^2 =
\operatorname{argmin}_\theta \| y - X \theta \|^2.
\end{equation}

@@

\begin{equation}
f'(\theta) = \sum_{i = 1}^n -2 x_i (y_i - x_i \theta) = -2 X^T (y - X \theta).
\end{equation}

@@

% TODO: syntax highlighting
\begin{verbatim}
class LinearRegressionFunction
{
 public:
  // Construct the LinearRegressionFunction with the given data.
  LinearRegressionFunction(arma::mat& X, arma::vec& y) : X(X), y(y) {}

  // Compute the objective function.
  double Evaluate(const arma::mat& theta)
  {
    return arma::accu((y - X * theta) % (y - X * theta));
  }

  // Compute the gradient and store in 'gradient'.
  void Gradient(const arma::mat& theta, arma::mat& gradient)
  {
    gradient = -2 * X.t() * (y - X * theta);
  }

 private:
  arma::mat& X;
  arma::vec& y;
};
\end{verbatim}

@@

\begin{verbatim}
LinearRegressionFunction lrf(X, y); // we assume X and y are loaded elsewhere
ens::L_BFGS lbfgs; // create L-BFGS optimizer with default parameters

arma::vec theta(X.n_rows, arma::fill::randu); // random uniform initialization
lbfgs.Optimize(lrf, theta); // after this call, theta holds the solution
\end{verbatim}

% TODO if space: elaborate on the example to add SGD-like optimizers.  The code
% is a little more complex but not much.

\section{Available optimizers}

@@

% I guess to save some space we should group them.
\begin{itemize}
  \item {\bf SGD variants:} Stochastic Gradient Descent (SGD), Stochastic
      Coordinate Descent (SCD), Parallel Stochastic Gradient Descent (Hogwild!),
      Stochastic Gradient Descent with Restarts (SGDR), SMORMS3, AdaGrad,
      AdaDelta, RMSProp, Adam, AdaMax

@@

  \item {\bf Objective functions:} Neural Networks, Logistic regression,
      Matrix completion, Neighborhood Components Analysis, Regularized SVD,
      Reinforcement learning, Softmax regression, Sparse autoencoders,
      Sparse SVM
\end{itemize}

@@

% not even sure this is the right term
\section{Static polymorphism for speed}

@@

%% TODO: code snippet

During the optimization, the {\tt rosen()} function will be called repeatedly.
However, since what was given to the {\tt minimize()} function is a function
pointer, this pointer must be dereferenced at each call.  In addition, some
optimizations such as inlining (and many optimizations enabled by inlining) are
not possible when using function pointers.  Because it is possible that an
individual optimization routine may need to compute the objective function,
gradient, and related quantities many times, the overhead of this lookup may be
non-negligible.  As one example, when training neural networks with SGD-like
optimizers on a dataset with $N$ points, the gradient must be repeatedly
computed on mini-batches of points of size $b$, and the training will pass over
the whole dataset $e$ times.  For typical values of $N \sim 100M$, $b = 64$, and
$e = 100$, this means the gradient computation function is being called almost
200 million times!

@@

\section{Automatic Creation of Functions}

It is often the case that when optimizing a function $f(x)$, the computation of
the value $f(x)$ and its derivative $f'(x)$ involve the computation of identical
intermediate results.  Consider the linear regression objective function from
earlier:

\begin{equation}
f(\theta) = \| y - X\theta \|_F^2.
\end{equation}

@@

Using template metaprogramming in {\tt ensmallen}, we provide an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual {\tt Evaluate()} and {\tt Gradient()} functions, a user
may simply write an {\tt EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  For our earlier
\texttt{LinearRegressionFunction}, we could use an implementation that only
computes $(y - X \theta)$ once:

\begin{verbatim}
double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient)
{
  const arma::rowvec v = (y - X * theta); // Cache result.
  gradient = -2 * X.t() * v;
  return arma::accu(v % v); // Take squared norm of v.
}
\end{verbatim}

This method could be provided {\it instead} of {\tt Evaluate()} and {\tt
Gradient()}.

In essence, given a {\tt FunctionType} class that implements either an {\tt
Evaluate()} and {\tt Gradient()} method or an {\tt EvaluateWithGradient()}
method, template metaprogramming techniques are used to detect which of these
methods exist.  Then, a wrapper class will use suitable mix-ins
in order to provide the `missing' functionality~\cite{smaragdakis2000mixin}.  If
{\tt EvaluateWithGradient()} does not exist, the following code is generated:

% TODO: code block
\begin{verbatim}
double EvaluateWithGradient(const arma::mat& coordinates, arma::mat& gradient)
{
  Gradient(coordinates, gradient);
  return Evaluate(coordinates);
}
\end{verbatim}

Similarly, if {\tt Evaluate()} or {\tt Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.  Since all of this code generation is done at compile-time and not
at runtime, the compiler is able to perform dead code elimination % TODO: check
on the unused result, and may be able to avoid calculations entirely.  In some
cases, then, the generated {\tt Evaluate()} or {\tt Gradient()} functions may be
equivalently fast to what would be hand-written!

@@

\section{Experiments}

@@

First, we consider the simple and popular Rosenbrock
function~\cite{Rosenbrock1960}:

\begin{equation}
f([x_1 x_2]) = 100 (x_2 - x_1^2)^2 + (1 - x_1^2).
\end{equation}

To demonstrate the added overhead of other frameworks, we choose to use
simulated annealing~\cite{kirkpatrick1983optimization},
a gradient-free optimizer that only uses the objective function itself.
Simulated annealing will call the objective function very many times; for each
simulation we limit the optimizer to $100,000$ objective evaluations.  We
compare four frameworks for this task:

\begin{itemize}
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item {\tt Optim.jl}'s simulated annealing implementation with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} from GNU Octave's {\tt optim} package~\cite{octave}
\footnote{Another option here might be {\tt simulannealbnd()} from MATLAB's
Global Optimization Toolkit.  However, no license was available for these
simulations.}
\end{itemize}

Table~\ref{tab:rosenbrock_results} shows the average runtime
across 10 trials for $100000$ iterations of simulated annealing with each
implementation.  Of these frameworks, only Julia and {\tt ensmallen} are able to
avoid the function pointer dereference and take advantage of inlining and
related optimizations---and this difference is strongly reflected in the
results.

\begin{table}[t]
\begin{center}
\begin{tabular}{cccc}
\toprule
{\tt ensmallen} & {\tt scipy} & {\tt Optim.jl} & {\tt samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
{\bf 0.004s} & 1.044s & 0.021s & 2.920s \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with different
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.}
\label{tab:rosenbrock_results}
\end{table}

@@

\begin{itemize}
  \item {\tt ensmallen}: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Optim.jl+Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient
  \item {\tt Optim.jl+ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy.optimize.fmin\_l\_bfgs\_b}: implements both objective and
gradient
  \item {\tt bfgsmin} from GNU Octave's {\tt optim} package
\end{itemize}

@@

\begin{table}
\begin{center}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d = 100$, $n = 1k$ & $d = 100$, $n = 10k$ & $d = 100$, $n =
100k$ & $d = 1k$, $n = 100k$ \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen} & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
{\tt Optim.jl+Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt Optim.jl+ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy.optimize.fmin\_l\_bfgs\_b} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for linear regression function with several different dataset
sizes.  All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\end{table}

@@

\section{Conclusion}

@@

The development of {\tt ensmallen} does not include just the authors named
here but also a long list of other contributors.  See \url{%TODO about page
} for more information.

>>>>

\documentclass{article}
\usepackage[preprint]{nips_2018}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage[newfloat]{minted}
\usepackage{caption}
\usepackage{parcolumns}
\usepackage{booktabs}

@@

\newcolumntype{R}[2]{%
  >{\adjustbox{angle=#1,lap=\width-(#2)}\bgroup}%
  l%
  <{\egroup}%
}
\newcommand*\rot{\multicolumn{1}{R{30}{2.0em}}}

\setlength{\textfloatsep}{0.1cm}

@@

% Alphabetical ordering?
\author{Shikhar Bhardwaj \\
Delhi Technological University \\
Delhi, India 110042 \\
\texttt{shikhar\_bt2k15@dtu.ac.in}
\And
Ryan R. Curtin \\
RelationalAI \\
Atlanta, GA, USA 30318 \\
\texttt{ryan@ratml.org}
\And
Marcus Edel \\
Free University of Berlin \\
Arnimallee 7, 14195 Berlin \\
\texttt{marcus.edel@fu-berlin.de}
\And
Yannis Mentekidis
% any affiliation/email?
\And
Conrad Sanderson \\
Data61, CSIRO \\
Australia
}

@@

\begin{abstract}
\vspace*{-0.3em}
%% the abstract below still needs more meat and sharpening
We present a fast and flexible C++ library for mathematical optimization of 
arbitrary user-supplied functions,
which can be applied to many machine learning problems.
The library provides many pre-built optimizers
(including numerous variants of SGD and Quasi-Newton optimizers)
as well as a flexible framework for implementing new optimizers and objective functions.
Implementation of a new optimizer requires only one method
and implementation of a new objective function requires at most four C++ functions. 
This can aid in the  quick implementation and prototyping of new machine learning algorithms.
In contrast to optimization approaches in existing frameworks
(such as Julia, Shogun, Vowpal Wabbit, SciPy),
the proposed Ensmallen library supports more optimizations,
more types of objective functions,
as well as seamlessly supporting user-defined objective functions and optimizers.
Empirical comparisons show that ...
TODO: briefly describe the speedups obtained in the experiments
\end{abstract}

@@

Mathematical optimization is the workhorse of virtually all machine learning
algorithms.  For a given objective function $f(\cdot)$
(which may have a special structure or constraints),
almost all machine learning problems can be boiled down
to the following optimization form:
%
%\vspace*{-0.2em}
\begin{equation}
\operatorname{argmin} f(x)
\end{equation}
\vspace*{-1.5em}

@@

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms.
We present {\tt ensmallen}, a C++ optimization toolkit
that contains a wide variety of optimization techniques for many types of
objective functions.  Through the use of C++ template
metaprogramming~\cite{Alexandrescu2001},
{\tt ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

Although there are many existing machine learning optimization toolkits, few
are able to take explicit advantage of metaprogramming based code optimizations,
and few offer robust support for various types of objective functions.
For instance, deep learning
libraries like Caffe~\cite{jia2014caffe},
PyTorch~\cite{paszke2017automatic},
and TensorFlow~\cite{abadi2016tensorflow}
each contain a variety of optimization techniques.  However, these techniques are
limited to stochastic gradient descent (SGD) and SGD-like optimizers that
operate on small batches of data points at a time.  Other machine learning
libraries, such as {\tt scikit-learn}~\cite{pedregosa2011scikit}
contain optimization algorithms but not in a coherent or reusable framework.
Many programming languages have higher-level packages for
mathematical optimization.  For example, {\tt
scipy.optimize}~\cite{jones2014scipy},
is widely used in the Python community, and MATLAB's function optimization
support has been available and used for many decades.  However, these
implementations are often unsuitable for modern machine learning tasks---for
instance, computing the full gradient of the objective function may not be
feasible because there are too many data points.

@@

% TODO: also, we should talk up the use

\vspace*{-0.2em}
\section{Types of objective functions}
\vspace*{-0.6em}

\begin{table}[b!]
\vspace*{-1.0em}

@@

        & & \multicolumn{7}{c}{} \\[0.6ex]

@@

\caption{\footnotesize{

@@

{\it unified framework} indicates if there is some kind of generic/unified

@@

{\it categorical} refers to if support for categorical features exists.
}}

@@

\vspace*{-1.8em}

@@

{\tt ensmallen} provides a {\bf set of optimizers} for
{\bf optimizing user-defined objective functions}.  It's also easy to implement a
new optimizer in the {\tt ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
%\vspace*{-0.4em}
%\begin{equation}
$\operatorname{argmin}_{x} f(x)$
%\end{equation}
%\vspace*{-0.4em}
%\noindent
for any function $f(x)$ that takes some vector or matrix input $x$.  Ideally,
the user should only need to specify what $f(x)$ is (or provide an
implementation of a function that can compute $f(x)$).  But it is impossible to
implement something both so generic and fast---for instance, gradient-based
approaches converge far more quickly than gradient-free approaches (in general),
so we must design an abstraction that is able to simultaneously generalize to
many problem types, as well as take advantage of accelerations and
optimizations.

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\bf arbitrary}: no assumptions can be made on $f(x)$
  \item {\bf differentiable}: $f(x)$ has a computable gradient $f'(x)$
  \item {\bf separable}: $f(x)$ is a sum of individual components: $f(x) =
\sum_{i} f_i(x)$
  \item {\bf categorical}: $x$ contains elements that can only take discrete
values
  %\item {\bf numeric}: all elements of $x$ take values in $\mathcal{R}$
  \item {\bf sparse}: the gradient $f'(x)$ or $f_i(x)$ (for a separable
function) is sparse
  \item {\bf partially differentiable}: the gradient $f'_j(x)$ is computable for
individual elements $x_j$ of $x$
  \item {\bf bounded}: $x$ is limited in the values that it can take
\end{itemize}
\vspace*{-0.3em}

@@

For the most common case of a differentiable $f(x)$, the user must
implement only two methods:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item \texttt{double Evaluate($x$)}: given coordinates $x$, this function
should return the value of $f(x)$.
  \item \texttt{void Gradient($x$, $g$)}: given coordinates $x$ and a reference
to $g$, set $g = f'(x)$.
\end{itemize}
\vspace*{-0.3em}

@@

\vspace*{-0.3em}
\section{Example: Learning Linear Regression Models}
\vspace*{-0.5em}

As an example of usage, consider the linear regression objective
function\footnote{Just for simplicity, we ignore the bias term.  It can be
rederived by taking $x^*_i = (x_i, 1)$.}.  Given some dataset $X \in
\mathcal{R}^{n \times d}$ and associated responses $y \in \mathcal{R}^n$, the
model of linear regression is to assume that $y_i = x_i \theta$ for each
point and response $(x_i, y_i)$.  To fit this model $\theta \in \mathcal{R}^d$ best
to the data, we must find

\vspace*{-1.1em}
\begin{equation}
\operatorname{argmin}_\theta \sum_{i = 1}^n (y_i - x_i \theta)^2 =
\operatorname{argmin}_\theta \| y - X \theta \|^2.
\end{equation}
\vspace*{-1.1em}

@@

\vspace*{-1.1em}
\begin{equation}
f'(\theta) = \sum_{i = 1}^n -2 x_i (y_i - x_i \theta) = -2 X^T (y - X \theta).
\end{equation}
\vspace*{-1.1em}

@@

% Indented a little to make it clear it's different from the text...
\vspace*{-0.4em}
\begin{minted}[fontsize=\small]{c++}
    class LinearRegressionFunction {
     public:
      // Construct the LinearRegressionFunction with the given data.
      LinearRegressionFunction(arma::mat& X, arma::vec& y) : X(X), y(y) {}

      // Compute the objective function.
      double Evaluate(const arma::mat& theta) {
        return arma::accu((y - X * theta) % (y - X * theta));
      }
      // Compute the gradient and store in 'gradient'.
      void Gradient(const arma::mat& theta, arma::mat& gradient) {
        gradient = -2 * X.t() * (y - X * theta);
      }
     private:
      arma::mat& X; arma::vec& y;
    };
\end{minted}
\vspace*{-0.4em}

@@

\vspace*{-0.4em}
\begin{minted}[fontsize=\small]{c++}
    LinearRegressionFunction lrf(X, y); // we assume X and y are loaded elsewhere
    ens::L_BFGS lbfgs; // create L-BFGS optimizer with default parameters

    arma::vec theta(X.n_rows, arma::fill::randu); // random uniform initialization
    lbfgs.Optimize(lrf, theta); // after this call, theta holds the solution
\end{minted}
\vspace*{-0.4em}

Only a slight variation on the signature of {\tt Evaluate()} and {\tt
Gradient()} would be needed to use the small-batch SGD-like optimizers that {\tt
ensmallen} provides.

\vspace*{-0.3em}
\section{Available optimizers}
\vspace*{-0.5em}

@@

% I guess to save some space we should group them.
\vspace*{-0.4em}
\begin{itemize}
  \item {\bf SGD variants:} Stochastic Gradient Descent (SGD), Stochastic
      Coordinate Descent (SCD), Parallel Stochastic Gradient Descent (Hogwild!),
      Stochastic Gradient Descent with Restarts (SGDR), SMORMS3, AdaGrad,
      AdaDelta, RMSProp, Adam, AdaMax

@@

% These were a part of mlpack but not ensmallen.
  %\item {\bf Objective functions:} Neural Networks, Logistic regression,
  %    Matrix completion, Neighborhood Components Analysis, Regularized SVD,
  %    Reinforcement learning, Softmax regression, Sparse autoencoders,
  %    Sparse SVM
\end{itemize}
\vspace*{-0.4em}

@@

% not even sure this is the right term
\vspace*{-0.3em}
\section{Static polymorphism for speed}
\vspace*{-0.5em}

@@

\vspace*{-0.3em}
\begin{minted}[fontsize=\footnotesize]{python}
    def rosen(x): return 100.0 * (x[1] - x[2]**2.0)**2.0 + (1 - x[2])**2.0)
    res = scipy.optimize.minimize(rosen, [-1.2, 1], method='nelder-mead')
\end{minted}
\vspace*{-0.3em}

During the call to {\tt minimize()}, the {\tt rosen()} function will be called
repeatedly---but it will be called as a function pointer, which must be
dereferenced each time {\tt rosen()} is called.  In addition, some
optimizations such as inlining (and many optimizations enabled by inlining) are
not possible when using function pointers.  Since some optimization algorithms
need to compute the objective very many times, the overhead of the lookup and
missed optimization opportunities may be non-negligible.  This applies even for
SGD-like optimizers that train machine learning models on very large datasets!

@@

\vspace*{-0.3em}
\section{Automatic generation of methods}
\vspace*{-0.5em}

It is often the case that when optimizing a function $f(x)$, the computation of
the value $f(x)$ and its derivative $f'(x)$ involve the computation of identical
intermediate results.  Consider the linear regression objective function from
earlier:
%\begin{equation}
$f(\theta) = \| y - X\theta \|_F^2.$
%\end{equation}

@@

Using template metaprogramming, {\tt ensmallen} provides an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual {\tt Evaluate()} and {\tt Gradient()} functions, a user
may simply write an {\tt EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  For our earlier
\texttt{LinearRegressionFunction}, we could use an implementation that only
computes $(y - X \theta)$ once:

\vspace*{-0.5em}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::rowvec v = (y - X * theta); // Cache result.
      gradient = -2 * X.t() * v;
      return arma::accu(v % v); // Take squared norm of v.
    }
\end{minted}
\vspace*{-0.5em}

% This is up here so that it's on the right page...
\begin{table}[t]
\begin{center}
\begin{tabular}{cccc}
\toprule
{\tt ensmallen} & {\tt scipy} & {\tt Optim.jl} & {\tt samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
{\bf 0.004s} & 1.044s & 0.021s & 2.920s \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with different
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.}
\label{tab:rosenbrock_results}
\end{table}

This method could be provided {\it instead} of {\tt Evaluate()} and {\tt
Gradient()}.  In either case, template metaprogramming techniques are used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide any `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if {\tt EvaluateWithGradient()} is not provided, a version will be
generated that calls both {\tt Gradient()} and {\tt Evaluate()} in turn.
Similarly, if {\tt Evaluate()} or {\tt Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

Since all of this code generation is done at compile-time and not
at runtime, the compiler is able to perform dead code elimination % TODO: check
on the unused result, and may be able to avoid calculations entirely.  In some
cases, then, the generated {\tt Evaluate()} or {\tt Gradient()} functions may be
equivalently fast to what would be hand-written!

@@

\vspace*{-0.3em}
\section{Experiments}
\vspace*{-0.5em}

@@

First, we consider the added overhead of other frameworks.  For our experiment,
we choose the simple and popular Rosenbrock function~\cite{Rosenbrock1960}:
$f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 - x_1^2)$.  For the optimizer, we use
simulated annealing~\cite{kirkpatrick1983optimization},
a gradient-free optimizer.  Simulated annealing will call the objective function
very many times; for each simulation we limit the optimizer to $100,000$
objective evaluations.  We compare four frameworks for this task:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item {\tt Optim.jl}'s simulated annealing implementation with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} from GNU Octave's {\tt optim} package~\cite{octave}
\footnote{Another option here might be {\tt simulannealbnd()} from MATLAB's
Global Optimization Toolkit.  However, no license was available for these
simulations.}
\end{itemize}
\vspace*{-0.3em}

Table~\ref{tab:rosenbrock_results} shows the average runtime
across 10 trials for $100000$ iterations of simulated annealing with each
implementation.  The simulations were run on a % TODO: what are Marcus's
% system specs?
{\tt \$COMPUTER}.
Of these frameworks, only Julia and {\tt ensmallen} are able to avoid the
function pointer dereference and take advantage of inlining and related
optimizations---and this difference is strongly reflected in the results.

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin} from GNU Octave's {\tt optim} package
\end{itemize}
\vspace*{-0.3em}

@@

\begin{table}
\begin{center}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d = 100$, $n = 1k$ & $d = 100$, $n = 10k$ & $d = 100$, $n =
100k$ & $d = 1k$, $n = 100k$ \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen} & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
{\tt Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy.optimize} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for linear regression function with several different dataset
sizes.  All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\end{table}

@@

\vspace*{-0.3em}
\section{Conclusion}
\vspace*{-0.5em}

@@

The development team of {\tt ensmallen} does not include just the authors named
here but also a long list of other contributors.  See \url{%TODO about page
} for more information.
